Report for Qualifying Exam

### 2024-08-14 src/sections/introduction.tex
added conclusion section, minor changes

This example clearly demonstrates how a term with a specific, legitimate meaning can be repurposed to spread misinformation and conspiracy theories.
The shift creates a potential data void where searches for \emph{crisis actor} might lead to a mix of factual information about emergency training and harmful conspiracy theories,
making it difficult for users to distinguish accurate information.
\\
This report introduces methodologies for analyzing data voids based on two key papers.
To contextualize our work, we present a thorough literature review examining the evolution of word meaning.
Subsequently, Section \Cref{sec:paper_kulkarni} delves into the first paper, focusing on significant linguistic change detection,
while Section \Cref{sec:paper_hamilton} explores the second paper on diachronic word embeddings.

>>>>

This example clearly demonstrates how a term with a specific, legitimate meaning can be repurposed to spread misinformation and conspiracy theories.
The shift creates a potential data void where searches for \emph{crisis actor} might lead to a mix of factual information about emergency training and harmful conspiracy theories,
making it difficult for users to distinguish accurate information.\\

This report introduces methodologies for analyzing data voids based on two key papers.
To contextualize our work, we present a thorough literature review examining the evolution of word meaning (\Cref{sec:relatedwork}).
Subsequently, Section \Cref{sec:paper_kulkarni} delves into the first paper, focusing on significant linguistic change detection,
while Section \Cref{sec:paper_hamilton} explores the second paper on diachronic word embeddings.

### 2024-08-14 src/sections/paper_hamilton.tex
added conclusion section, minor changes

\begin{table}[tbh]
\begin{tabular}{@{}llll@{}}
\toprule
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}}                                          & \multicolumn{3}{c}{\cellcolor[HTML]{FFFFFF}\textbf{Examples}}                                                                                                                                                                                                                               \\ \cmidrule(l){2-4}
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\multirow{-2}{*}{\cellcolor[HTML]{FFFFFF}\textbf{Semantic Shift}}} & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{PPMI}}                                                    & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SVD}}                  & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SGNS}}                                        \\ \midrule
\rowcolor[HTML]{EFEFEF}
\textbf{Genuine Shift}                                                                & started                                                                                                      & \begin{tabular}[c]{@{}l@{}}headed, calls,\\ gay, actually\end{tabular}    & \begin{tabular}[c]{@{}l@{}}wanting, gay, check, starting,\\ major, actually, headed\end{tabular} \\
\cellcolor[HTML]{FFFFFF}\textbf{Borderline}                                           & \begin{tabular}[c]{@{}l@{}}know, got, would, \\ decided, think, stop, \\ remember, must, wanted\end{tabular} & male, naturally                                                           & touching                                                                                         \\
\rowcolor[HTML]{EFEFEF}
\textbf{Corpus Artifact}                                                              & -                                                                                                             & \begin{tabular}[c]{@{}l@{}}harry, wherever,\\ special, cover\end{tabular} & harry, romance                                                                                   \\ \bottomrule
\end{tabular}
\caption{The top 10 words identified by each embedding method were examined, and after reviewing the literature and analyzing their nearest neighbors,
the authors categorized these words into different types of semantic shifts.
For instance, the word “headed” shifted from primarily meaning the “top of a body or entity” to signifying “a direction of travel” indicated a \emph{genuine} shift.
Additionally, some \emph{borderline} cases were identified, which were largely attributed to global shifts in genre or discourse, as well as corpus artifacts.
For example, words like “special,” “cover,” and “romance” appeared to be \emph{artifacts} from the covers of fiction books, occasionally influenced by advertisements,
taken from \cite{hamilton-etal-2016-diachronic}}
\label{tab:discovering-shifts-table}
\end{table}

@@

\subsection{Takeaways}\label{subsec:takeaways4}
This paper explores how different types of word embedding models, such as PPMI, SVD, and SGNS, can be used to study diachronic shifts in word meanings.
By analyzing these aligned embeddings, the authors identify two key statistical laws of semantic change.
Law of conformity shows that high-frequency words have smaller semantic displacements over time.
They used statistical models to assess the relationship between word frequency and the rate of semantic change. Specifically, they found that the logarithm of a word's frequency had a significant negative effect on the rates of semantic change, indicating that higher frequency words change less rapidly
The Law of Innovation is established by demonstrating that polysemous words—those with diverse meanings—tend to experience greater semantic change. The clustering coefficient was used as a key metric for measuring polysemy within co-occurrence networks, revealing that words with lower clustering coefficients (higher polysemy) are more prone to semantic shifts. This law highlights the dynamic nature of words with multiple meanings and their tendency to innovate in the face of linguistic evolution.

>>>>

\begin{table}[tbh]
\centering
\small
\begin{tabular}{@{}llll@{}}
\toprule
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}}                                          & \multicolumn{3}{c}{\cellcolor[HTML]{FFFFFF}\textbf{Examples}}                                                                                                                                                                                                                               \\ \cmidrule(l){2-4}
\rowcolor[HTML]{FFFFFF}
\multicolumn{1}{c}{\multirow{-2}{*}{\cellcolor[HTML]{FFFFFF}\textbf{Semantic Shift}}} & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{PPMI}}                                                    & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SVD}}                  & \multicolumn{1}{c}{\cellcolor[HTML]{FFFFFF}\textbf{SGNS}}                                        \\ \midrule
\rowcolor[HTML]{EFEFEF}
\textbf{Genuine Shift}                                                                & started                                                                                                      & \begin{tabular}[c]{@{}l@{}}headed, calls,\\ gay, actually\end{tabular}    & \begin{tabular}[c]{@{}l@{}}wanting, gay, check, starting,\\ major, actually, headed\end{tabular} \\
\cellcolor[HTML]{FFFFFF}\textbf{Borderline}                                           & \begin{tabular}[c]{@{}l@{}}know, got, would, \\ decided, think, stop, \\ remember, must, wanted\end{tabular} & male, naturally                                                           & touching                                                                                         \\
\rowcolor[HTML]{EFEFEF}
\textbf{Corpus Artifact}                                                              & -                                                                                                             & \begin{tabular}[c]{@{}l@{}}harry, wherever,\\ special, cover\end{tabular} & harry, romance                                                                                   \\ \bottomrule
\end{tabular}
\caption{The top 10 words identified by each embedding method were examined, and after reviewing the literature and analyzing their nearest neighbors,
the authors categorized these words into different types of semantic shifts.
For instance, the word “headed” shifted from primarily meaning the “top of a body or entity” to signifying “a direction of travel” indicated a \emph{genuine} shift.
Additionally, some \emph{borderline} cases were identified, which were largely attributed to global shifts in genre or discourse, as well as corpus artifacts.
For example, words like “special,” “cover,” and “romance” appeared to be \emph{artifacts} from the covers of fiction books, occasionally influenced by advertisements,
taken from \cite{hamilton-etal-2016-diachronic}.}
\label{tab:discovering-shifts-table}
\end{table}

@@

\subsection{Takeaways}\label{subsec:takeaways4}
This paper explores how different types of word embedding models, such as PPMI, SVD, and SGNS, can be used to study diachronic shifts in word meanings.
By analyzing these aligned embeddings, the authors identify two key statistical laws of semantic change.
Law of conformity shows that high-frequency words have smaller semantic displacements over time.
They used statistical models to assess the relationship between word frequency and the rate of semantic change. Specifically, they found that the logarithm of a word's frequency had a significant negative effect on the rates of semantic change, indicating that higher frequency words change less rapidly
The Law of Innovation is established by demonstrating that polysemous words—those with diverse meanings—tend to experience greater semantic change. The clustering coefficient was used as a key metric for measuring polysemy within co-occurrence networks, revealing that words with lower clustering coefficients (higher polysemy) are more prone to semantic shifts.
This law highlights the dynamic nature of words with multiple meanings and their tendency to innovate in linguistic evolution.

### 2024-08-14 src/sections/relatedwork.tex
added conclusion section, minor changes

\para{Semantic shifts with contextual embeddings:}
Giulianelli et al.\ (2020) explored the phenomenon of lexical semantic change using an unsupervised approach using \emph{BERT}.
They utilized the \emph{BERT} model to generate contextual word embeddings, which captured the meaning of a word based on its surrounding context.
The extracted word usage vectors are then clustered into different \emph{usage types} using the \emph{k-means clustering algorithm}.
This helped to identify distinct senses or meanings of the words as they appear in various contexts.
They analyzed these clusters for a specific word across different time periods.
The study proposed three metrics to quantify semantic change:
\begin{packed_enumerate}
    \item \textbf{Entropy Difference (ED)}: Measures the change in uncertainty (entropy) of a word’s usage distribution over time.
    \item \textbf{Jensen-Shannon Divergence (JSD)}: Compares the similarity of word usage distributions across time intervals.
    \item \textbf{Average Pairwise Distance (APD)}: Computes the average distance between word usage vectors from different periods, indicating shifts in word meaning.
\end{packed_enumerate}
The qualitative analysis indicated that the approach could capture various linguistic phenomena, including both synchronic (current usage) and diachronic (historical changes) aspects.

@@

% ------------------------------------------------------------------ SYNTACTIC ------------------------------------------------------------------
\subsection{Syntactic Alterations} \label{subsec:syntactic-alterations}
\para{Linguistic change studies have shifted to include \emph{syntax} alongside semantics.}
The field of linguistic change has traditionally focused on \emph{semantics}.
However, recent studies have begun to explore the role of \emph{syntax} in language evolution as well.
The \emph{syntactic} functionality of a word can evolve by transitioning into a new part-of-speech (POS) category.
Nouns, due to their inherent flexibility in meaning, exhibit a greater tendency to undergo these changes driven by cultural shifts.
While verbs are more likely to participate in gradual semantic changes that follow established linguistic patterns. \cite{hamilton-etal-2016-cultural, hamilton-etal-2016-diachronic}

@@

% ------------------------------------------------------------------ THREE PAPERS ------------------------------------------------------------------
\subsection{Selected Papers}\label{subsec:selected-papers}
Several studies from our literature review provide valuable insights for addressing data voids in our analysis.
The first being the work by Kulkarni et al.\ which introduces us to time series construction.
Their distributional methods focuses on finding subtle semantic shits to determine the context where a word in used.
This concept aligns perfectly with our goal of uncovering shifts in word usage when encountering data voids.

Second paper by Hamilton et al.\ introduces us to word emebeddings alignment.
Since we want to compare word vectors from different time periods, vectors should be aligned in same coordinate axes.
After aligning the embeddings for individual time periods, we can use the aligned word vectors to compute the semantic displacement that a word has undergone during a certain time-period.
Next two section \Cref{sec:paper_kulkarni}, \Cref{sec:paper_hamilton} focusses on the selected papers in detail.

>>>>

\para{Semantic shifts with contextual embeddings:}
Giulianelli et al.\ (2020) explored the phenomenon of lexical semantic change using an unsupervised approach using \emph{BERT}.
They utilized the \emph{BERT} model to generate contextual word embeddings, which captured the meaning of a word based on its surrounding context.
The extracted word usage vectors are then clustered into different \emph{usage types} using the \emph{k-means clustering algorithm}.
This helped to identify distinct senses or meanings of the words as they appear in various contexts.
They analyzed these clusters for a specific word across different time periods.
The study proposed three metrics to quantify semantic change:\\
\textbf{Entropy Difference (ED)}: Measures the change in uncertainty (entropy) of a word’s usage distribution over time.\\
\textbf{Jensen-Shannon Divergence (JSD)}: Compares the similarity of word usage distributions across time intervals.\\
\textbf{Average Pairwise Distance (APD)}: Computes the average distance between word usage vectors from different periods, indicating shifts in word meaning.

The qualitative analysis indicated that the approach could capture various linguistic phenomena, including both synchronic (current usage) and diachronic (historical changes) aspects.

@@

% ------------------------------------------------------------------ SYNTACTIC ------------------------------------------------------------------
\subsection{Syntactic Alterations} \label{subsec:syntactic-alterations}
\para{Linguistic change studies have shifted to include \emph{syntax} alongside semantics.}
The field of linguistic change has traditionally focused on \emph{semantics}.
However, recent studies have begun to explore the role of \emph{syntax} in language evolution as well.
The \emph{syntactic} functionality of a word can evolve by transitioning into a new part-of-speech (POS) category.
Nouns, due to their inherent flexibility in meaning, exhibit a greater tendency to undergo these changes driven by cultural shifts.
While verbs are more likely to participate in gradual semantic changes. \cite{hamilton-etal-2016-cultural, hamilton-etal-2016-diachronic}

@@

% ------------------------------------------------------------------ THREE PAPERS ------------------------------------------------------------------
\subsection{Selected Papers}\label{subsec:selected-papers}
Several studies from our literature review provide valuable insights for addressing data voids in our analysis.
The first being the work by Kulkarni et al.\ which introduces us to time series construction.
Their distributional methods focuses on finding subtle semantic shits to determine the context where a word in used.
This concept aligns perfectly with our goal of uncovering shifts in word usage when encountering data voids.
Second paper by Hamilton et al.\ introduces us to word emebeddings alignment.
Since we want to compare word vectors from different time periods, vectors should be aligned in same coordinate axes.
After aligning the embeddings for individual time periods, we can use the aligned word vectors to compute the semantic displacement that a word has undergone during a certain time-period.
Next two section \Cref{sec:paper_kulkarni}, \Cref{sec:paper_hamilton} focusses on the selected papers in detail.
